---
title: "Untitled"
author: "Xiaoting Chen"
date: "2023-11-13"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**report maximum pvalue for sensitivity analysis**


## 1. 

Consider question 1 from Homework 3 about the effect of being physically active on
living longer. Use one of your matchings from Homework 3 (e.g., your pair matching) to
perform a sensitivity analysis.

### (a) 
Up to what value of is there still evidence that being physically active causes
you to live longer (under the significance level 0.05)?

```{r}
# if choose fullmatching
library(sensitivityfull)
```

## 2. 
Dynarski (2003, American Economic Review) asked the question: does offering
financial aid to college students affect college attendance? During the period 1979-1981,
the Social Security Student Benefit provided substantial college tuition aid for a high
school senior whose father had died. Dynarski compared college attendance for high
school seniors who had a deceased father. The data are in the file educationaid7981.csv
on the course website. The potential confounders we would like to control for are family
income (faminc), income missing (incmiss), black, hispanic, the Armed Forces
Qualification Test (afqtpct), whether mother attended college (momcoll), whether father
attended college (fathercoll) and gender (female)1
. The treatment indicator f_dead is 1 for
a senior whose father is deceased and 0 for other seniors. The outcome is attend.college
(1 if the person has attended college by age 23, 0 if not).

### (a) 
Construct a match with adequate balance and test whether having a deceased
father increases college attendance under the assumption of no unmeasured
confounding (you could consider several matches and choose the one you think is
best, but make sure to choose the match before testing the outcome to preserve the
blinding advantages of matching).

```{r}

```

### (b) 
Conduct a sensitivity analysis to address concerns about unmeasured
confounding.
```{r}

```


